TrueGrid: Code the Table, Tabulate the Data

### 2016-06-16 true-grid/paper_SEMS16.tex
comments by reviewers addressed

In this paper we propose a method to bring both worlds closer together, by juxtaposing ordinary code and spreadsheet-like grids in the IDE, called TrueGrid. Using TrueGrid spreadsheet cells can be programmed using full featured programming language. Spreadsheet users then may enjoy benefits of source code, including added abstractions, syntax highlighting, version control, etc. On the other hand, programmers may leverage the grid for interactive exploring and testing of code. 
We illustrate these benefits using a prototype implementation of TrueGrid that runs in the browser an uses Javascript as a programming language.
\end{abstract}

@@

This liveness is one of the core features of most spreadsheet systems. When a users enters a formula and presses enter, they see the result, without long edit-compile-run cycles. This live characteristic of spreadsheets is often praised as their key success factor~\cite{HermansPhD}.
\begin{figure}
  \begin{center}
  \includegraphics[width=9cm]{fig/bret.png}
  \caption{Live programming: on the right the source code and on the left its instantiation of the code which changes immediately when the code is updated, screenshot from~\cite{Victor2012}.}
  \label{fig:bret}
  \end{center}
\end{figure} 
However, spreadsheets have a number of downsides too. For instance, the lack of  abstraction mechanisms forces spreadsheet users to use copy-paste for reusing code. Although solutions to this problem have been researched~\cite{hermans_2015_19341,jones2003user}, they do not provide the power of full programming languages to spreadsheet users. Secondly, the way formulas are edited in a spreadsheet system like Excel has important drawbacks from the programming perspective. For instance, Excel's formula editor lacks even the basic editor services, such as  syntax highlighting and reference resolution.  In general, the interface is not inviting to apply proper coding styles and practices. Another disadvantage of embedding the code within the sheet itself is that it prohibits versioning and sharing of the formulas separately from the data.

@@

In this paper, we will describe \textit{TrueGrid}, a light-weight approach to bridging the gap between programming and spreadsheets and describe ts implications in both worlds. The basic concept of TrueGrid is to allow a spreadsheet like grid to be programmed using a full featured programming language, in a consistent user interface. Figure \ref{fig:TG} shows this idea as implemented in our prototype. 
The key charactertistic is that  developers see the code and data at once. Furthermore, like a spreadsheet, TrueGrid is live, i.e. on a change of data or code, the grid is updated.
In our example, we use JavaScript as a language, however, the idea itself is not limited to a single programming language.
% in fact, one can even image a grid in which different programming languages live.

@@

As an implementation of TrueGrid for spreadsheet users, we hypothesize that existing spreadsheet system is extended with a source code editor view, next to the grid-based view of the data. This allows users to  use advanced features like PivotTables and charts on top of their TrueGrid, but also use professional IDE services for expressing computations. For spreadsheet users, using TrueGrid over spreadsheets presents several benefits. For example, the use of a professional  editor supports the developer with editor services like syntax highlighting and error marking. Furthermore, the textual form of the code allows for easy diffing and merging, enabling more mature version control on spreadsheets. While learning a new programming language can be challenging, there are spreadsheet developers working with VBA now, which is a fully featured language. Unfortunately the integration between the code and a spreadsheet is low level and cumbersome. We envision TrueGrid having a lower lower learning curve because code and data are juxtaposed.

@@

As an example consider the simple grade book sheet shown in Fig.~\ref{FIG:grades}. It shows the actual data (both provided and computed) in the grid. The average and class average are expressed using ordinary Javascript functions. The TrueGrid environment links functions or methods to the cells in the grid using a naming convention.
For instance, function starting with the \lstinline{column_} prefix computes a complete column, given a particular row (an ordinary Javascript object).
This is used in the computation of the \lstinline{avg} column, where each cell contains the average of the lab and exam cells.
The model also allows naming individual cells. This is illustrated in the function \lstinline{cell_classAvg}. The function receives the current column (\lstinline{col}, a Javascript array), and the current \lstinline{row}. Based on the elements in the column the class average is computed.

@@

In particular, TrueGrid can be seen as a persistent REPL, where expressions or method invocations are continuously evaluated, after every change to the code or input data. 
Consider the example shown in Fig.~\ref{FIG:leftpad}.
On the left is a simple Javascript function for left-padding values with spaces or other padding characters. 
Using TrueGrid, the programmer can provide example data in the grid and explore the implementation. This can be especially valuable early in the development process when  you have some data and only a vague idea of where the how certain functionality should be implemented.

@@

Spreadsheets are live programming environments. However, their lack of abstraction mechanisms and editor support are impediments to professional spreadsheet development. Conversely, traditional programming environments lack the continuos feedback that make spreadsheets so attractive. In this paper we have presented TrueGrid, a user interface design combining code editing and grid-based data view, with a live execution model.

TrueGrid presents a promising bridge between the domain of spreadsheets and software development. It has the potential to improve end-user programming experience from two perspectives:
\begin{itemize}
\item For spreadsheet users: computations are expressed using program code, instead of  formulas in cells, so that end-users may enjoy both abstraction and liveness at the same time.
\item For professional developers: spreadsheet-like grids support live exploration and testing of code, without explicitly invoking test scripts of entering expressions in a REPL.
\end{itemize}
We have implemented a prototype of TrueGrid which runs in the browser, using Javascript as the programming language. Further research is needed to empirically investigate the benefits of TrueGrid from both the programmer and spreadsheet user perspectives. 
We expect that TrueGrid provides a fruitful vehicle for exploring the middle ground between full end-user programming, and professional software development.

>>>>

In this paper we propose a method to bring both worlds closer together, by juxtaposing ordinary code and spreadsheet-like grids in the IDE, called TrueGrid. Using TrueGrid spreadsheet cells can be programmed with a fully featured programming language. Spreadsheet users then may enjoy benefits of source code, including added abstractions, syntax highlighting, version control, etc. On the other hand, programmers may leverage the grid for interactive exploring and testing of code. 
We illustrate these benefits using a prototype implementation of TrueGrid that runs in the browser and uses Javascript as a programming language.
\end{abstract}

@@

This liveness is one of the core features of most spreadsheet systems. When a users enters a formula and presses enter, they see the result, without a lengthy edit-compile-run cycle. This live characteristic of spreadsheets is often praised as their key success factor~\cite{HermansPhD}.
\begin{figure}
  \begin{center}
  \includegraphics[width=9cm]{fig/bret.png}
  \caption{Live programming: on the right the source code and on the left its instantiation of the code which changes immediately when the code is updated, screenshot from~\cite{Victor2012}.}
  \label{fig:bret}
  \end{center}
\end{figure} 
However, spreadsheets have a number of downsides too. For instance, the lack of  abstraction mechanisms forces spreadsheet users to use copy-paste for reusing code. Although solutions to this problem have been researched~\cite{hermans_2015_19341,jones2003user}, they do not provide the power of full programming languages to spreadsheet users. Secondly, the way formulas are edited in a spreadsheet system like Excel has important drawbacks from the programming perspective. For instance, Excel's formula editor lacks even the basic editor services, such as  syntax highlighting and reference resolution.  In general, the interface is not inviting to apply proper coding styles and practices. Another disadvantage of embedding the code within the sheet itself is that it prohibits versioning and sharing of the formulas separately from the data.

@@

In this paper, we will describe \textit{TrueGrid}, a light-weight approach to bridging the gap between programming and spreadsheets and describe ts implications in both worlds. The basic concept of TrueGrid is to allow a spreadsheet like grid to be programmed using a full featured programming language, in a consistent user interface. Figure \ref{fig:TG} shows this idea as implemented in our prototype. 
The key charactertistic is that  developers see the code and data at the same time. Furthermore, like a spreadsheet, TrueGrid is live, i.e. on a change of data or code, the grid is updated.
In our example, we use JavaScript as a language, however, the idea itself is not limited to a single programming language.
% in fact, one can even image a grid in which different programming languages live.

@@

As an implementation of TrueGrid for spreadsheet users, we hypothesize that existing spreadsheet system could be extended with a source code editor view, next to the grid-based view of the data. This allows users to  use advanced features like PivotTables and charts on top of their TrueGrid, but also use professional IDE services for expressing computations. For spreadsheet users, using TrueGrid over spreadsheets presents several benefits. For example, the use of a professional  editor supports the developer with editor services like syntax highlighting and error marking. Furthermore, the textual form of the code allows for easy diffing and merging, enabling more mature version control on spreadsheets. While learning a new programming language can be challenging, there are spreadsheet developers working with VBA now, which is a fully featured language. Unfortunately the integration between the code and a spreadsheet is low level and cumbersome. We envision TrueGrid having a less steep learning curve, because code and data are juxtaposed.

@@

As an example consider the simple grade book sheet shown in Fig.~\ref{FIG:grades}. It shows the actual data (both provided and computed) in the grid. The average and class average are expressed using ordinary Javascript functions. The TrueGrid environment links functions or methods to the cells in the grid using a naming convention.
For instance, a function starting with the \lstinline{column_} prefix computes a complete column, given a particular row (an ordinary Javascript object).
This is used in the computation of the \lstinline{avg} column, where each cell contains the average of the lab and exam cells.
The model also allows naming individual cells. This is illustrated in the function \lstinline{cell_classAvg}. The function receives the current column (\lstinline{col}, a Javascript array), and the current \lstinline{row}. Based on the elements in the column the class average is computed.

@@

In particular, TrueGrid can be seen as a persistent REPL, where expressions or method invocations are continuously evaluated, after every change to the code or input data. 
Consider the example shown in Fig.~\ref{FIG:leftpad}.
On the left is a simple Javascript function for left-padding values with spaces or other padding characters. 
Using TrueGrid, the programmer can provide example data in the grid and explore the implementation. This can be especially valuable early in the development process when  you have some data and only a vague idea of how certain functionality should be implemented.

@@

Spreadsheets are live programming environments. However, their lack of abstraction mechanisms and editor support are impediments to professional spreadsheet development. Conversely, traditional programming environments lack the continuous feedback that makes spreadsheets so attractive. In this paper we have presented TrueGrid, a user interface design combining code editing and grid-based data view, with a live execution model.

TrueGrid presents a promising bridge between the domain of spreadsheets and software development. It has the potential to improve end-user programming experience from two perspectives:
\begin{itemize}
\item For spreadsheet users: computations are expressed using program code, instead of  formulas in cells, so that end-users may enjoy both abstraction and liveness at the same time.
\item For professional developers: spreadsheet-like grids support live exploration and testing of code, without explicitly invoking test scripts of entering expressions in a REPL.
\end{itemize}
We have implemented a prototype of TrueGrid which runs in the browser, using Javascript as the programming language. Further research is needed to empirically investigate the benefits of TrueGrid from both the programmer and spreadsheet user perspectives. 
We expect that TrueGrid provides a fruitful vehicle for exploring the middle ground between end-user programming on the one hand side, and professional software development on the other.
